---
title: "Lorenz Equations"
author: "Stephen Gaito"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Lorenz Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE,  results="asis"}
options(rgl.useNULL=TRUE)
library(rgl)
library(rglwidget)
library(dynSysToolKit)

setupKnitr() # contains an implicit rgl.open()?

```

## Model equations

We investigate the [Lorenz Equations](https://en.wikipedia.org/wiki/Lorenz_system) using the RGL graphics tool. A good [A complete guide to 3D visualization device system in R - R software and data visualization](http://www.sthda.com/english/wiki/a-complete-guide-to-3d-visualization-device-system-in-r-r-software-and-data-visualization#install-the-rgl-package) provides an excellent tutorial. To learn how to adjust the mouse buttons see: [Translation using mouse for rgl plot](http://stackoverflow.com/questions/23996006/translation-using-mouse-for-rgl-plot)

Consider using the tools [deSolve](https://cran.r-project.org/web/packages/deSolve/index.html) and [rootSolve](https://cran.r-project.org/web/packages/rootSolve/index.html)

$$
\begin{align*}
\frac{dx}{dt} & = & \sigma ( y - x ) \\
\frac{dy}{dt} & = & x ( \rho - z ) - y \\
\frac{dz}{dt} & = & x y - \beta z
\end{align*}
$$

Where interesting values for sigma, rho and beta are:

$$
\begin{align*}
\sigma & = & 10 \\
\rho   & = & 28 \\
\beta  & = & \frac{8}{3}
\end{align*}
$$

```{r lorenzEqs, webgl=TRUE, rgl.newwindow=TRUE, echo=FALSE}
shortResults <- integrateLorenz(stepsPerSample = 1, numSamples = 100)
results <- integrateLorenz(stepsPerSample = 1, numSamples = 10000)
#rgl.linestrips(results[,1], results[,2], results[,3], color = "blue")
x <- results[,1]
y <- results[,2]
z <- results[,3]
#plot3d(x, y, z, type='l', col="blue")
lorenzID <- lines3d(x,y,z, col="blue", radius=0.1)
shortID  <- lines3d(shortResults[,1], shortResults[,2], shortResults[,3], col="red")
axisID   <- decorate3d()
subid    <- currentSubscene3d()
#rglwidget(elementId = "lorenzEqs")
```

<script type="text/javascript">
var myTranslate = function(xDist, yDist) {
  var rgl = document.getElementById("lorenzEqsdiv").rglinstance;
  var zoom = rgl.getObj(`r subid`).par3d.zoom;
  rgl.getObj(`r subid`).par3d.userMatrix.translate( xDist*zoom, yDist*zoom, 0);
  rgl.drawScene();
};
</script>

<button type="button" onclick="myTranslate(-10, 0)">Left</button>
<button type="button" onclick="myTranslate(10, 0)">Right</button>
<button type="button" onclick="myTranslate(0, 10)">Up</button>
<button type="button" onclick="myTranslate(0, -10)">Down</button>

```{r results="asis", echo=FALSE}
toggleButton(lorenzID, label = "full",    prefix = "lorenzEqs", subscene = subid)
toggleButton(shortID,  label = "initial", prefix = "lorenzEqs", subscene = subid)
toggleButton(axisID,   label = "axis",    prefix = "lorenzEqs", subscene = subid)
```
