---
title: "Specifying a model in R"
author: "Stephen Gaito"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying a model in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Specifying a model in R

```{r echo=FALSE}
library(ecoNetworksSimulator)
```

To build a given trophic model we begin by creating an empty trophic model and then adding the different species:

```{r}
r2c1 <- newTrophicModel()
r2c1 <- addSpecies(r2c1, 
  "resource1",
  growthRate=0.2,
  carryingCapacity=25,
  timeLag = 0.0)
r2c1 <- addSpecies(r2c1, 
  "resource2",
  growthRate=0.2,
  carryingCapacity=25,
  timeLag = 3.5)
r2c1 <- addSpecies(r2c1, 
  "consumer",
  mortality=0.2,
  halfSaturation=1)
r2c1
```

Next we add some interactions:

```{r}
r2c1 <- addInteraction(r2c1,
  predator="consumer",
  prey="resource1",
  attackRate=0.1,
  conversionRate=0.35,
  deathRate = 1.0,
  timeLag = 0.0)
r2c1 <- addInteraction(r2c1,
  predator="consumer",
  prey="resource2",
  attackRate=0.1,
  conversionRate=0.35,
  deathRate = 1.0,
  timeLag = 2.3)
r2c1
```

we can then get a LaTeX representation of our model

```{r eval=FALSE}
getLaTeX(r2c1) # NOT YET IMPLEMENTED
```

or the species interactions in an interactions *matrix* form

```{r eval=FALSE}
getInteractionsMatrix(r2c1) # NOT YET IMPLEMENTED
```

We might want to specify a random collection of models based upon individual species and interaction mean and standard deviations. To do this we can add species standard deviations tables to the model:

```{r}
r2c1 <- addSpeciesStd(r2c1, 
  name="resource1",
  growthRate=0.01,
  carryingCapacity=5,
  timeLag = 0.01)
r2c1 <- addSpecies(r2c1, 
  name="resource2",
  growthRate=0.01,
  carryingCapacity=5,
  timeLag = 0.1)
r2c1 <- addSpecies(r2c1, 
  name="consumer",
  mortality=0.01,
  halfSaturation=0.0)
r2c1
```

Next we add some interactions standard deviations:

```{r}
r2c1 <- addInteractionStd(r2c1,
  predator="consumer",
  prey="resource1",
  attackRate=0.01,
  conversionRate=0.1,
  deathRate = 0.01,
  timeLag = 0.01)
r2c1 <- addInteractionStd(r2c1,
  predator="consumer",
  prey="resource2",
  attackRate=0.01,
  conversionRate=0.1,
  deathRate = 0.01,
  timeLag = 0.1)
r2c1
```

To get an individual model:

```{r eval=FALSE}
# r2c1 <- varyModel(r2c1var) # NOT YET IMPLEMENTED
```
